Optimal updating factor in adaptive flat-distribution-sampling simulations

### 2016-04-21 doc/invt.tex
new routine to compute optimal q(T)
invt.tex: rewrite characterization of the optimal schedule

%For the single-bin scheme
%used in WL algorithm,
Later,
it was found that
a better way
of controlling the schedule $\alpha(t)$
in late stages
is to follow the formula
%
\begin{equation}
  \alpha(t) = \frac{1}{t},
  \label{eq:alpha_invt}
\end{equation}
%
where $t$,
referred to as the ``time'' below,
is the number of steps
from the beginning of the simulation.
%
This simple yet effective formula has attracted
several studies\cite{belardinelli2007, belardinelli2007jcp, belardinelli2008,
morozov2007, zhou2008, morozov2009,
komura2012, caparica2012, caparica2014}.
%
There are, however, some reservation about
the constant of proportionality:
the optimal formula for $\alpha(t)$
might be $C/t$ with a $C$
different from $1.0$\cite{
morozov2007, zhou2008, morozov2009}.

@@

As shown above, with a fixed value of $q(T)$,
the optimization of the total error is equivalent to
that of the asymptotic part.
%
We can rewrite Eq. \eqref{eq:error_asym1} as an action
%
\begin{equation}
  \Err_A
  =
  \int_0^T
    {\mathcal L} \bigl[ q(t)\bigr]
    \, dt
  ,
%\notag
\label{eq:error_asym_Lagrangian}
\end{equation}
%
where the Lagrangian,
%
\begin{align}
  {\mathcal L} \bigl[ q(t) \bigr]
  &=
  \sum_{ k = 2 }^n
    \Gamma_k \, \dot u_k^2\bigl[ q(t) \bigr]
  \notag
  \\
  &=
  \sum_{ k = 2 }^n
    \Gamma_k \, \lambda_k^2 \, u_k^2 \bigl[ q(t) \bigr]
  \; \dot q^2( t ),
\notag
\end{align}
%
represents a free particle
with a position-dependent mass.
%
Since the Lagrangian
does not explicitly depend on time, $t$,
the Hamiltonian
%
\begin{equation}
  \mathcal H
  =
  \frac{ \partial \mathcal L }
       { \partial \dot q     }
  \, \dot q
  -
  \mathcal L
  =
  2 \, \mathcal L
  - \mathcal L
  =
  \mathcal L,
  %=
  %\mathrm{const.}
  \notag
  %\label{eq:error_asym_Hamiltonian}
\end{equation}
%
which coincides with the Lagrangian,
is conserved.
%
So we may set
%
\begin{equation}
  \mathcal L
  =
    \textstyle\sum_{ k = 2 }^n
      \Gamma_k \, \lambda_k^2
      \, u_k^2 \bigl[ q(t) \bigr]
  \;
  \dot q^2(t)
  =
  c^2
  ,
  \quad
  (c > 0).
  \label{eq:Lagrangian_const}
\end{equation}
%
Integrating the square root, we get
%
\begin{equation}
  \int_{ 0 }^{ q(t) }
    \sqrt{
      \textstyle\sum_{ k = 2 }^n
        \Gamma_k \, \lambda_k^2
        \, u_k^2( q' )
    }
    \;
    d q'
  =
  c \, t
  ,
  \label{eq:q_opt}
\end{equation}
%
where $c$ is to be determined by
the boundary condition at $t = T$ as
%
\begin{equation}
  c =
  \frac 1 T
  \int_{ 0 }^{ q(T) }
    \sqrt{
      \textstyle\sum_{ k = 2 }^n
        \Gamma_k \, \lambda_k^2
        \, u_k^2( q' )
    }
    \;
    d q'
  .
  \notag
\end{equation}
%
Thus, we have an implicit equation for $q(t)$,
and hence the optimal schedule,
$\alpha(t) = \dot q(t)$.
%
Since we only optimized the asymptotic error, $E_A$,
in th above,
the parameter $q(T)$ can be further adjusted
to minimize the total error, $E = E_R + E_A$.
%
A complete program of computing the optimal schedule
is listed in Appendix \ref{sec:program}.

\subsubsection{\label{sec:mass_distr}
Characterization of the optimal schedule}

We can characterize the optimal schedule
geometrically as follows.
%
First, we define a mass function
$$
  M(Q)
  =
  \sqrt{
    \textstyle\sum_{ k = 2 }^n \Gamma_k \, \lambda_k^2 \, e^{-2 \, \lambda_k \, Q}
  }
  .
$$
and a normalized mass distribution as
\begin{equation}
  m(Q)
  =
  \frac{
    M(Q)
  }
  {
    \int_0^{ q(T) } M(Q') \, d Q'
  }
  ,
\label{eq:mass_distr}
\end{equation}
%
such that
$\int_0^{q(T)} m(Q) \, dQ = 1$.
%
As shown in Fig. \ref{fig:massq},
the distribution peaks at $Q = 0$,
and decreases with $Q$.
%
According to Eq. \eqref{eq:q_opt},
the ordinate of the distribution curve
gives $1/(T\alpha) = 1/(T \dot q)$,
and the area under the curve in the region $Q' < Q$
gives $t/T$.
%
From this picture, we can draw several conclusions.

@@

For the asymptotic error, we have,
from Eqs.
\eqref{eq:error_asym_Lagrangian}
and
\eqref{eq:Lagrangian_const},
%
\begin{equation}
  E_A = c^2 \, T
  =
  \frac 1 T
  \left(
    \int_0^{ q(T) } M(Q) \, dQ
  \right)^2
  .
\label{eq:error_asym2}
\end{equation}
%
The optimal value of $q(T)$ that minimizes
the total error satisfies
\begin{equation}
  m\bigl( q(T) \bigr)
  =
  \frac{1} { T \, \alpha_0 / 2 }
  ,
\label{eq:opt_qT}
\end{equation}
i.e., the initial updating magnitude
should be half of $\alpha_0$.

\note{Let $Q_c \equiv q(T)$,
$$
\begin{aligned}
  \frac{
    \partial E_R
  }
  {
    \partial q(T)
  }
  &=
  -\alpha_0 \, M^2(Q_c)
  ,
  \\
  \frac{
    \partial E_A
  }
  {
    \partial q(T)
  }
  &=
  \frac 2 T \,
  M(Q_c) \,
  \int_0^{ Q_c } M(Q) \, dQ
  .
\end{aligned}
$$
Then solve $\partial (E_R + E_A) / \partial q(T) = 0$.
}

First,
the optimal schedule has the following scaling property.
%
For two simulations of lengths $T_1$ and $T_2$,
if $q(T_1) = q(T_2)$
and they share the same $\lambda_k$'s and $\Gamma_k$'s,
the updating magnitudes at the corresponding times,
$t_1/T_1 = t_2/T_2$,
can be related as $T_1 \, \alpha_1 = T_2 \, \alpha_2$.
%
In other words,
if $\alpha(t_1)$ is
an optimal schedule of the simulation of length $T_1$,
then
$\frac{T_1}{T_2} \alpha\left( \frac{ T_1 } { T_2 } t_2 \right)$
is an optimal schedule of the simulation of length $T_2$.
%
The asymptotic errors are related by
$T_1 \, E_{A1} = T_2 \, E_{A2}$.

@@

In the above, we have focused on
the asymptotic error.
%
For a finite-length simulation,
we should include the residual part
and minimize the total error.
%
For the bandpass updating schemes,
this leads to the optimal value
of the free parameter, $t_0$.
%
%Assuming the system is initially equilibrated
%under a constant updating magnitude,
%$\alpha_0$,
Assuming Eq. \eqref{eq:error_res1} with
$$
q(T)
=
\int_0^T \frac{ dt } { t + t_0 }
=
\ln \frac{ T + t_0 } { t_0 }.
$$
%
we have, from Eqs. \eqref{eq:error_split}
and \eqref{eq:error_asym_sinc},
%
\begin{equation}
  E
  =
  E_R + E_A
  =
  \frac { \frac 1 2 \alpha_0 \, t_0^2 + T }
        {          ( T + t_0 )^2          }
  \sum_{ k = 2 }^K \Gamma_k
  .
\notag
%\label{eq:error_sinc0}
\end{equation}
%
The minimum is reached at
%
\begin{equation}
  t_0
  =
  \frac{     2    }
       { \alpha_0 }
  ,
\label{eq:t0_sinc}
\end{equation}
%
and at this point
%
\begin{equation}
  E
  =
  \frac{   1     }
       { T + t_0 }
  \sum_{ k = 2 }^K
    \Gamma_k
  .
\label{eq:error_sinc}
\end{equation}
%
In Appendix \ref{sec:equilerr},
we show that the above error tends
to the error of the ideal equilibrium FDS simulation
in the asymptotic limit.

Note also that at the beginning of the schedule,
the updating magnitude is
$\alpha(0^+) = \alpha_0 / 2$,
half of the previous equilibrium value.
%
This factor $1/2$
happens to coincide with the
reduction factor of updating magnitude
at stage transitions
in the original WL algorithm\cite{
wang2001, wang2001pre}.

@@

As an example of the multiple-bin schemes,
we chose the nearest-neighbor updating scheme
described in Sec. \ref{sec:nnscheme}.
%
We choose $\mu_1 = 0.24$ to the smallest eigenvalue
close to zero.
%
Again, simulation and theory agreed well,
as shown in Fig. \ref{fig:errnnbr}.
%
The normalized error
and the optimal value of $1/\lambda$,
depended on the simulation length, $T$,
in the global sampling case,
because of the near zero eigenvalues,
as discussed in Appendix \ref{sec:invt_schedule}.
%
%The optimal $1/\lambda$ gradually increased
%with simulation length,
%as discussed in Sec. \ref{sec:optlambda}.
%
However, for the local sampling case,
the $T$-dependence was suppressed
because of the dominance of large $\Gamma_k$'s
of the small-$k$ modes.
%
The optimal $\lambda$ remained around $1.0$,
as suggested by Eq. \eqref{eq:lambda_nn_onestep}.

@@

\item
Adjust $q(T)$,
to satisfy Eq. \eqref{eq:opt_qT}.

@@

\section{\label{sec:equilerr}
Comparison to the error of the equilibrium FDS simulation}

@@

The error of the equilibrium FDS method
depends on the precision of the histogram.
%
If all $v_i$'s are exact,
the error is minimized, and can be computed
from the second term
on the right-hand side of Eq. \eqref{eq:vcorr_equil},
%
\begin{align}
  E
  &=
  \left\langle
    \sum_{ i = 1 }^ n
      p_i \,
      \left(
        \ln \frac { H_i }
                  { p_i }
      \right)^2
  \right\rangle
%\notag
%\\
  %&
  \approx
  \sum_{ i = 1 }^ n
    p_i \,
    \left\langle
      \left(
        \frac { H_i }
              { p_i }
        - 1
      \right)^2
    \right\rangle,
\notag
\end{align}
where
$H_i = \frac{1}{T} \sum_{t = 1}^T h_i(t)$
with
$h_i(t)$ defined by Eq. \eqref{eq:h_def}.

>>>>

%For the single-bin scheme
%used in WL algorithm,
Later,
it was found that
a better way
of controlling the schedule $\alpha(t)$
in late stages
is to follow the formula
%
\begin{equation}
  \alpha(t) = \frac{1}{t},
  \label{eq:alpha_invt}
\end{equation}
%
where $t$,
referred to as the ``time'' below,
is the number of steps
from the beginning of the simulation.
%
This simple yet effective formula has attracted
several studies\cite{belardinelli2007, belardinelli2007jcp, belardinelli2008,
morozov2007, zhou2008, morozov2009,
komura2012, caparica2012, caparica2014}.
%
There are, however, some reservation about
the constant of proportionality:
the optimal formula for $\alpha(t)$
might be $c/t$ with a $c$
different from $1.0$\cite{
morozov2007, zhou2008, morozov2009}.

@@

As shown above, with a fixed value of $q(T)$,
the optimization of the total error is equivalent to
that of the asymptotic part.
%
We can rewrite Eq. \eqref{eq:error_asym1} as an action
%
\begin{equation}
  \Err_A
  =
  \int_0^T
    {\mathcal L} \bigl[ q(t)\bigr]
    \, dt
  ,
%\notag
\label{eq:error_asym_Lagrangian}
\end{equation}
%
where the Lagrangian,
%
\begin{align}
  {\mathcal L} \bigl[ q(t) \bigr]
  &=
  \sum_{ k = 2 }^n
    \Gamma_k \, \dot u_k^2\bigl[ q(t) \bigr]
  \notag
  \\
  &=
  \sum_{ k = 2 }^n
    \Gamma_k \, \lambda_k^2 \, u_k^2 \bigl[ q(t) \bigr]
  \; \dot q^2( t ),
\notag
\end{align}
%
represents a free particle
with a position-dependent mass.
%
Since the Lagrangian
does not explicitly depend on time, $t$,
the Hamiltonian
%
\begin{equation}
  \mathcal H
  =
  \frac{ \partial \mathcal L }
       { \partial \dot q     }
  \, \dot q
  -
  \mathcal L
  =
  2 \, \mathcal L
  - \mathcal L
  =
  \mathcal L,
  %=
  %\mathrm{const.}
  \notag
  %\label{eq:error_asym_Hamiltonian}
\end{equation}
%
which coincides with the Lagrangian,
is conserved.
%
So we may set
%
\begin{equation}
  \mathcal L
  =
    \sum_{ k = 2 }^n
      \Gamma_k \, \lambda_k^2
      \, u_k^2 \bigl[ q(t) \bigr]
  \;
  \dot q^2(t)
  =
  \left(
    \frac C T
  \right)^2
  ,
  \label{eq:Lagrangian_const}
\end{equation}
%
for some positive $C$.
%
Integrating the square root, we get
%
\begin{equation}
  \int_{ 0 }^{ q(T) }
    \sqrt{
      \textstyle\sum_{ k = 2 }^n
        \Gamma_k \, \lambda_k^2
        \, u_k^2( q' )
    }
    \;
    d q'
  =
  C \, \frac t T
  ,
  \label{eq:q_opt}
\end{equation}
%
where $c$ is to be determined by
the boundary condition at $t = T$ as
%
\begin{equation}
  C =
  \int_{ 0 }^{ q(t) }
    \sqrt{
      \textstyle\sum_{ k = 2 }^n
        \Gamma_k \, \lambda_k^2
        \, u_k^2( q' )
    }
    \;
    d q'
  .
  \notag
\end{equation}
%
Thus, we have an implicit equation for $q(t)$,
and hence the optimal schedule,
$\alpha(t) = \dot q(t)$.
%
Since we only optimized the asymptotic error, $E_A$,
in th above,
the parameter $q(T)$ can be further adjusted
to minimize the total error, $E = E_R + E_A$.
%
A complete program of computing the optimal schedule
is listed in Appendix \ref{sec:program}.

\subsubsection{\label{sec:mass_distr}
Characterization of optimal schedules}

We can characterize the optimal schedule
geometrically as follows.
%
First, we define the mass function as
$$
  M(Q)
  =
  \sqrt{
    \textstyle\sum_{ k = 2 }^n \Gamma_k \, \lambda_k^2 \, e^{-2 \, \lambda_k \, Q}
  }
  .
$$
and the normalized mass distribution as
\begin{equation}
  m(Q)
  =
  \frac{
    M(Q)
  }
  {
    \int_0^{ q(T) } M(Q') \, d Q'
  }
  =
  \frac{
    M(Q)
  }
  {
    C
  }
  ,
\label{eq:mass_distr}
\end{equation}
%
such that
$\int_0^{q(T)} m(Q) \, dQ = 1$.
%
Clearly,
the distribution decreases monotonically with $Q$.

@@

Then, from Eq. \eqref{eq:q_opt},
we find that for $Q = q(T) - q(t)$,
%
\begin{align}
  m(Q)
  &=
  \frac{ 1 }
       { T \, \alpha(t) }
  ,
\label{eq:mQ_invTa}
  \\
  \int_Q^{ q(T) }
    m(Q') \, dQ'
  &=
  \frac t T
  ,
\label{eq:intmQ_tT}
\end{align}
%
which means that each point $(t, \alpha(t))$
on the schedule can be mapped
to a point $(Q, m(Q))$ on the mass distribution,
such that the ordinate of the latter curve
gives $1/(T\alpha)$,
and the area under the curve in the domain $[Q, q(T)]$
is equal to $t/T$,
as shown in Fig. \ref{fig:massq},

This mapping allows us to classify all optimal schedules
as follows.
%
We collect in each class optimal schedules
computed from simulations with the same function $M(Q)$
and cutoff value $q(T)$.
%
The former is determined by the values of
$\lambda_k$'s and $\Gamma_k$'s,
whereas the latter affects the distribution $m(Q)$
via the normalization constant $C$.

In each class, the updating magnitudes of two schedules
of lengths $T_1$ and $T_2$
at corresponding times $t_1/T_1 = t_2/T_2$
are related by $T_1 \, \alpha_1 = T_2 \, \alpha_2$.
%
Further, by Eqs.
\eqref{eq:error_asym_Lagrangian}
and
\eqref{eq:Lagrangian_const},
we have
%
\begin{equation}
  E_A
  =
  \frac { C^2 } { T }
  =
  \frac 1 T
  \left(
    \int_0^{ q(T) } M(Q) \, dQ
  \right)^2
  ,
\label{eq:error_asym2}
\end{equation}
%
which means the asymptotic error
is inversely proportional to the length
for the same-class schedules,
and we have the scaling relation
$T_1 \, E_{A1} = T_2 \, E_{A2}$.

%The optimal schedule has the following scaling property.
%%
%For two simulations of lengths $T_1$ and $T_2$,
%if $q(T_1) = q(T_2)$
%and they share the same $\lambda_k$'s and $\Gamma_k$'s,
%the updating magnitudes at the corresponding times,
%$t_1/T_1 = t_2/T_2$,
%can be related as $T_1 \, \alpha_1 = T_2 \, \alpha_2$.
%%
%In other words,
%if $\alpha(t_1)$ is
%an optimal schedule of the simulation of length $T_1$,
%then
%$\frac{T_1}{T_2} \alpha\left( \frac{ T_1 } { T_2 } t_2 \right)$
%is an optimal schedule of the simulation of length $T_2$.
%%
%The asymptotic errors are related by
%$T_1 \, E_{A1} = T_2 \, E_{A2}$.

\subsubsection{\label{sec:optinitalpha}
  Optimal initial updating magnitude
}

Let us now consider simulations in which
the system is initially equilibrated at the same
constant updating magnitude $\alpha$.
%
For these simulations,
we can adjust $q(T)$
to minimize the total error.
%
From Eqs. \eqref{eq:error_res1}
and \eqref{eq:error_asym2},
we find that
the $q(T)$ that minimizes
the total error satisfies
\begin{equation}
  m\bigl( q(T) \bigr)
  =
  \frac{1} { T \, (\alpha_0 / 2) }
  .
\label{eq:opt_qT}
\end{equation}
%
\note{Let $Q_c \equiv q(T)$,
$$
\begin{aligned}
  \frac{
    \partial E_R
  }
  {
    \partial q(T)
  }
  &=
  -\alpha_0 \, M^2(Q_c)
  ,
  \\
  \frac{
    \partial E_A
  }
  {
    \partial q(T)
  }
  &=
  \frac 2 T \,
  M(Q_c) \,
  \int_0^{ Q_c } M(Q) \, dQ
  .
\end{aligned}
$$
Then solve $\partial (E_R + E_A) / \partial q(T) = 0$.
}
%
Then, by Eqs. \eqref{eq:mQ_invTa} and \eqref{eq:intmQ_tT},
we have
\begin{equation}
  \alpha( t = 0 )
  =
  \frac{ \alpha_0 }
       { 2 }
  ,
\label{eq:half_alpha0}
\end{equation}
%
i.e., the optimal initial updating magnitude
is always half of the equilibration value.
%
This factor $1/2$
happens to coincide with the
reduction factor of updating magnitude
at stage transitions
in the original WL algorithm\cite{
wang2001, wang2001pre}.

Besides,
for simulations with the same $\alpha_0$,
Eq. \eqref{eq:opt_qT} shows that
the optimal $q(T)$ increases with
the simulation length $T$.
%
If $M(Q)$ decays rapidly,
the distribution $m(Q)$ does not critically
depend on the cutoff $q(T)$,
as long as $q(T)$ is sufficiently large.
%
So the optimal schedules of these simulations
approximately satisfy the scaling relations
described in Sec. \ref{sec:mass_distr}.
%
If, however, $M(Q)$ has a long tail
because of some near-zero $\lambda_k$'s,
increasing the simulation length $T$
generally alters the shape of the optimal schedule.

@@

In the above, we have focused on
the asymptotic error.
%
For a finite-length simulation,
we should include the residual part
and minimize the total error.
%
For the bandpass updating schemes,
this leads to the optimal value
of the free parameter, $t_0$,
which is, according to Eq. \eqref{eq:half_alpha0}
%
%%Assuming the system is initially equilibrated
%%under a constant updating magnitude,
%%$\alpha_0$,
%Assuming Eq. \eqref{eq:error_res1} with
%$$
%q(T)
%=
%\int_0^T \frac{ dt } { t + t_0 }
%=
%\ln \frac{ T + t_0 } { t_0 }.
%$$
%%
%we have, from Eqs. \eqref{eq:error_split}
%and \eqref{eq:error_asym_sinc},
%%
%\begin{equation}
%  E
%  =
%  E_R + E_A
%  =
%  \frac { \frac 1 2 \alpha_0 \, t_0^2 + T }
%        {          ( T + t_0 )^2          }
%  \sum_{ k = 2 }^K \Gamma_k
%  .
%\notag
%%\label{eq:error_sinc0}
%\end{equation}
%%
%The minimum is reached at
%
\begin{equation}
  t_0
  =
  \frac{     2    }
       { \alpha_0 }
  ,
\label{eq:t0_sinc}
\end{equation}
%
and at this point
%
\begin{equation}
  E
  =
  \frac{   1     }
       { T + t_0 }
  \sum_{ k = 2 }^K
    \Gamma_k
  .
\label{eq:error_sinc}
\end{equation}
%
In Appendix \ref{sec:equilerr},
we show that the above error tends
to the error of the ideal equilibrium FDS simulation
in the asymptotic limit.

@@

As an example of the multiple-bin schemes,
we chose the nearest-neighbor updating scheme
described in Sec. \ref{sec:nnscheme}.
%
We choose $\mu_1 = 0.24$ to the smallest eigenvalue
close to zero.
%
Again, simulation and theory agreed well,
as shown in Fig. \ref{fig:errnnbr}.
%
The normalized error
and the optimal value of $1/\lambda$,
depended on the simulation length, $T$,
in the global sampling case,
because of the near-zero eigenvalues,
as discussed in Appendix \ref{sec:invt_schedule}.
%
%The optimal $1/\lambda$ gradually increased
%with simulation length,
%as discussed in Sec. \ref{sec:optlambda}.
%
However, for the local sampling case,
the $T$-dependence was suppressed
because of the dominance of large $\Gamma_k$'s
of the small-$k$ modes.
%
The optimal $\lambda$ remained around $1.0$,
as suggested by Eq. \eqref{eq:lambda_nn_onestep}.

@@

\item
Compute $q(T)$ from the root of Eq. \eqref{eq:opt_qT}.

@@

\section{\label{sec:equilerr}
Error of the ideal equilibrium FDS simulation}

@@

The error of the equilibrium FDS simulation
depends on the precision of the histogram.
%
If all $v_i$'s are exact,
the error is minimized, and can be computed
from the second term
on the right-hand side of Eq. \eqref{eq:vcorr_equil},
%
\begin{align}
  E
  &=
  \left\langle
    \sum_{ i = 1 }^ n
      p_i \,
      \left(
        \ln \frac { H_i }
                  { p_i }
      \right)^2
  \right\rangle
%\notag
%\\
  %&
  \approx
  \sum_{ i = 1 }^ n
    p_i \,
    \left\langle
      \left(
        \frac { H_i }
              { p_i }
        - 1
      \right)^2
    \right\rangle,
\notag
\end{align}
where
$H_i = \frac{1}{T} \sum_{t = 1}^T h_i(t)$
with
$h_i(t)$ defined by Eq. \eqref{eq:h_def}.
